---
title: "Make source data grids"
description: >
  This vignette shows how to build the netCDF data file and then
  update it to new start and end dates.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Make source data grids}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

```{r setup}
library(AWAPer, warn.conflicts = FALSE)
```
This example shows how to build the required data file and then update it.

## Make netCDF file
This example shows the steps required to build the netCDF data grid file.

First, let's define the dates for the data grids. Here the data grids are created for data between the dates _updateTo_ and _updateTo_. If the latter two dates were not input then data would be downloaded from 1/1/1900 to yesterday.

Note, in practice users often run _makeNetCDF_file_ once to build the netCDF data file that contain all variables over the entire record length (which requires ~5GB disk storage) and then use the netCDF grids for multiple projects, rather than re-building the netCDF grids for each project.

```{r}
startDate <- as.Date(Sys.Date()-15,"%Y-%m-%d")
endDate <- as.Date(Sys.Date()-5,"%Y-%m-%d")
```

Next the file name for the netCDF grids need to be defined. Here we'll just use a temporary file. You should change this to a non-temporary file name and folder.
```{r}
ncdfFilename <- tempfile(fileext='.nc')
```

Now we're ready to download and build the netCDF grids.

The netCDF data file contains grids of daily rainfall, minimum and maximum temperature. Below, this will be updated to include vapour pressure and solar radiation for all of Australia.
```{r}
ncdffile.name <- makeNetCDF_file(ncdfFilename=ncdfFilename,
                updateFrom=startDate, updateTo=endDate,
                vars = c('precip','tmin','tmax'))
```

Now let's get a summary of the netCDF file that we've created. Note that the output data.frame shows the expected variables over the expected date range.
```{r}
summary.df <- AWAPer::summary(ncdffile.name)
summary.df
```

## Add a variable to an existing data grids

Now that we've built the above file, the updating can be demonstrated. Here we'll update the data grids to also include vapour pressure. Here the same date range will be used. Because the date range of the new variable equals the date range of the existing variables, only vapour pressure data is updated. This feature can be used to update the data of a single variable as required.

```{r}
ncdffile.name <- makeNetCDF_file(ncdfFilename=ncdffile.name,
                updateFrom=startDate, updateTo=endDate,
                vars = c('vprp'))
```

Now let's check that the file includes both the original three variable plus vapour pressure.
```{r}
summary.df <- AWAPer::summary(ncdffile.name)
summary.df
```
## Add a variable to an existing data grids and update other variable

Now lets add solar radiation. Here an earlier start date will be used. However, because the date range of all variables within the netCDF file must be equal, the date range of the other variables must also be updated.

```{r}
startDate <- startDate - 5
```

```{r}
ncdffile.name <- makeNetCDF_file(ncdfFilename=ncdffile.name,
                updateFrom=startDate, updateTo=endDate,
                vars = c('solarrad'))
```

Now let's check that the file includes both the prior four three variable plus solar radiation.
```{r}
summary.df <- AWAPer::summary(ncdffile.name)
summary.df
```
