---
title: "Extract daily point precipitation"
description: >
  This vignette shows how to build the netCDF data file and then
  extract the point precipitation data at a few locations over a usr defined period.
  The extracted data is then compared with precipitation gauge data.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extract daily point precipitation}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

```{r setup}
library(AWAPer)
```
This example shows how to build the required data files and then extract the daily rainfall at two sites. Here, one of the sites is at a rainfall gauge and the results are compared against observations.

## Make netCDF file
The first step is to create the netCDF files. Here the netCDF file is created and only for precipitation data and only between the dates _dataFrom_ and _dataTo_. If the latter two dates were not input then data would be downloaded from 1/1/1900 to yesterday.
```{r}
dataFrom = as.Date("2010-08-01","%Y-%m-%d")
dataTo = as.Date("2010-10-01","%Y-%m-%d")
```

The netCDF file contains grids of daily rainfall for all of Australia and is used below to extract data at points of interest. Often users run _makeNetCDF_file_ once to build netCDF data files that contain all variables over the entire record length (which requires ~5GB disk storage) and then use the netCDFs grids for multiple projects, rather than re-building the netCDF for each project. Also, if _makeNetCDF_file_ is run with the the netCDF file name pointing to existing files and _updateFrom=NA_ then the netCDF files will be updated to yesterday.

To start to create the grids, the file name for the netCDF grids needs to be defined. Here we'll just use a temporary file. You should change this to a non-temporary file name and folder.
```{r}
ncdfFilename = tempfile(fileext='.nc')
```
Now with the inputs defined we can download the gridded data and build the netCDF file. Note, here _vars_ is set to only download daily precipitation.
```{r}
fnames = makeNetCDF_file(ncdfFilename = ncdfFilename,
                         updateFrom = dataFrom,
                         updateTo = dataTo,
                         vars = c('precip'))
```

## Set the points for data extraction
with the data grids built, let's set two data points to extract data at. The locations are at a groundwater bore and one rainfall gauge. Below, the coordinates are defined and then they're convert to a spatial object and set projection to GDA94.
```{r}
coordinates.data = data.frame( ID =c('Bore-10084446','Rain-63005'),
                               Longitude = c(153.551875, 149.5559),
                               Latitude =  c(-28.517974,-33.4289))

sp::coordinates(coordinates.data) <- ~Longitude + Latitude

sp::proj4string(coordinates.data) = '+proj=longlat +ellps=GRS80 +no_defs'
```

## Extract daily precipitation data
Now let's extract the daily precipitation at the two locations. The data is extracted from the netCDF file _ncdfFilename_ and between the dates _extractFrom_ and _extractTo_.

Note the netCDF file _ncdfFilename_ must be in the working directory or the full file path must be given.
```{r}
extracted.data = extractCatchmentData(ncdfFilename = ncdfFilename,
                                        extractFrom = dataFrom,
                                        extractTo = dataTo,
                                        locations = coordinates.data,
                                        vars = c('precip'),
                                        ET.function='')
```

## Plot the daily precipitation at each site
Now let's plot time series of the extracted daily precipitation.

```{r}
#| fig.width = 6,
#| fig.height = 6
for (i in 1:nrow(coordinates.data)){
  filt = extracted.data$Location.ID == coordinates.data$ID[i]

  data2plot = extracted.data$precip[filt]
  dates = ISOdate(extracted.data$year[filt],
                  extracted.data$month[filt],
                  extracted.data$day[filt])

  if (i==1){
    plot(dates, data2plot, main='Extracted precip.',
         ylab='Precip [mm/d]', xlab='Date [day/month]',
         type='l',lty = i+1, cex=0.2)
  } else {
    lines(dates, data2plot, lty = i+1)
  }
}
legend('topright',legend=c('Bore-10084446','Rain-63005'),lty=2:3)
```

## Check extracted data against precipitation gauge date
To assess the reliability of the extracted rain gauge data, let's use the following observed precipitation (from gauge_63005) sourced from the Australian Bureau of Meteorology.
```{r}
obsPrecip <- data.frame(
  year= c(2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010,
          2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010,
          2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010, 2010),
  month = c(8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 9, 9, 9, 9, 9,
            9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 9, 10),
  day = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 1, 2, 3, 4, 5,
          6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 1),
  precip = c(0.6, 5.2, 0.8, 0.4, 0.0, 0.0, 0.0, 0.0, 0.0, 15.8, 15.6, 7.6, 0.7, 0.4, 1.4, 1.0, 0.0, 0.0, 30.4, 1.0, 0.0,
                0.0, 5.0, 2.2, 0.3, 0.8, 13.8, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.5, 13.8, 15.2, 0.4, 0.2, 0.0, 0.0, 12.4, 0.9,
                0.0, 0.0, 0.1, 13.0, 0.2, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.3, 0.0, 0.0, 0.0, 0.0))
```
Now let's compare the extracted rainfall against the gauge rainfall data.

The left plot shows that the gauge and extracted data plot along the 1:1. This shows that the extracted data is unbiased. The scatter does, however, show that there is some difference in the estimates. This arises because the source grids are at a 0.05 degree resolution, or about 5 km x 5 km, and this introduces some error with the gauge is not at the centre of the grid cell and the location to be extracted is not also at the centre.

```{r}
#| fig.width = 8,
#| fig.height = 4
par(mfrow = c(1, 2))

filt2 = extracted.data$Location.ID=='Rain-63005'
plot(obsPrecip$precip,extracted.data$precip[filt2],
     xlim = c(0,35),ylim = c(0,35),
     main='Obs. vs. Extracted precip.',
     xlab='Obs. [mm/d]', ylab='Extracted precip. [mm/d]', cex=1)

abline(0,1, col='grey', lty=2)

plot(cumsum(obsPrecip$precip),cumsum(extracted.data$precip[filt2]),
     xlim = c(0,175),ylim = c(0,175),
     main='Cum. obs. vs. Extracted precip.',
     xlab='Obs. [mm]', ylab='Extracted precip. [mm]', type='l', cex=1)
abline(0,1, col='grey', lty=2)
```
