---
title: "Area weighted evapotranspiration measures"
description: >
  This vignette shows how to estimate various measures of the area weighted evapotranspiration.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Area weighted evapotranspiration measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

```{r setup}
library(BOMcatchr)
```

This example calculates and plot various estimates of evaportranspiration. Ten different estimates of area weighted evapotranspiration over one year at catchment 407214 (Victoria, Australia) are derived.

## Make netCDF files
Like the other vignettes, the netCDF data grids need to be built.

First, let's define the start and end dates for data grids and the file names.

```{r}
date.from = as.Date("2010-01-01","%Y-%m-%d")
date.to = as.Date("2010-12-31","%Y-%m-%d")

ncdfFilename = tempfile(fileext='.nc')
```

Next, let's make the data grids over this period.
```{r}
build.grids(ncdfFilename = ncdfFilename,
                   updateFrom = date.from,
                   updateTo = date.to,
                   vars = c('precip','tmin', 'tmax',
                   'vprp_3pm', 'solarrad'))
```

## Load catchment boundaries
Now that we have the meteorological data we can begin extracting data for two catchments. To make it easy to get the require location data, the package includes the function \code{extract_locations}, which pulls vector spatial data from the BOM Australian Hydrological Geospatial Fabric.  Here it to used to obtain the boundaries for two catchments within the Loddon basin, Victoria. However, shape files of polygons can easily be used once imported into R.
```{r}
catch <- extract_locations('catchment',c('407214','407220'))
site_ID <- catch$stationno
```

## Extract daily precipitation and PET data
Next, the 11 different measures of evapotranspiration that can be derived from the available gridded data are calculated for 12 months.

The estimation of ET uses the _evapotranspiration_ package. It requires a set of constants, which are loaded as follows.
```{r}
data(constants,package='Evapotranspiration')
```

Next, all 11 ET measures are derived. For each measure, only the following commands change: \code{ET.function} , \code{ET.timestep} and \code{ET.Mortons.est} (when Morton's estimate is derived).
```{R}
climateData.ET.HargreavesSamani = extract_data(ncdfFilename = ncdfFilename,
                                extractFrom = date.from,
                                extractTo = date.to,
                                vars = c('tmax', 'tmin', 'et'),
                                locations = catch,
                                spatial.fn = 'IQR',
                                ET.function = 'ET.HargreavesSamani',
                                ET.timestep = 'daily',
                                ET.constants = constants)

climateData.ET.JensenHaise = extract_data(ncdfFilename = ncdfFilename,
                                extractFrom = date.from,
                                extractTo = date.to,
                                vars = c('tmax', 'tmin', 'solarrad', 'et'),
                                locations = catch,
                                spatial.fn = 'IQR',
                                ET.function = 'ET.JensenHaise',
                                ET.timestep = 'daily',
                                ET.constants = constants)

climateData.ET.Makkink = extract_data(ncdfFilename = ncdfFilename,
                                extractFrom = date.from,
                                extractTo = date.to,
                                vars = c('tmax', 'tmin','solarrad', 'et'),
                                locations = catch,
                                spatial.fn = 'IQR',
                                ET.function = 'ET.Makkink',
                                ET.timestep = 'daily',
                                ET.constants = constants)

climateData.ET.McGuinnessBordne = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.McGuinnessBordne',
                               ET.timestep = 'daily',
                               ET.constants = constants)

climateData.ET.MortonCRAE = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'precip', 'vprp_3pm', 'solarrad', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.MortonCRAE',
                               ET.timestep = 'monthly',
                               ET.constants = constants)

climateData.ET.MortonCRAE.potentialET = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'precip', 'vprp_3pm', 'solarrad', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.MortonCRAE',
                               ET.timestep = 'monthly',
                               ET.Mortons.est = 'potential ET',
                               ET.constants = constants)

climateData.ET.MortonCRAE.wetarealET = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'precip', 'vprp_3pm', 'solarrad', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.MortonCRAE',
                               ET.timestep = 'monthly',
                               ET.Mortons.est = 'wet areal ET',
                               ET.constants = constants)

climateData.ET.MortonCRAE.actualarealET = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'precip', 'vprp_3pm', 'solarrad', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.MortonCRAE',
                               ET.timestep = 'monthly',
                               ET.Mortons.est = 'actual areal ET',
                               ET.constants = constants)

climateData.ET.MortonCRWE = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'precip', 'vprp_3pm', 'solarrad', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.MortonCRWE',
                               ET.timestep = 'monthly',
                               ET.Mortons.est = 'potential ET',
                               ET.constants = constants)

climateData.ET.MortonCRWE.shallowLake = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'precip', 'vprp_3pm', 'solarrad', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.MortonCRWE',
                               ET.timestep = 'monthly',
                               ET.Mortons.est = 'shallow lake ET',
                               ET.constants = constants)

climateData.ET.Turc = extract_data(ncdfFilename = ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('tmax', 'tmin', 'solarrad', 'et'),
                               locations = catch,
                               spatial.fn = 'IQR',
                               ET.function = 'ET.Turc',
                               ET.timestep = 'daily',
                               ET.constants = constants)
```
Next let's plot each measure of the area weighted evaporation at each catchment.

```{r}
#| fig.width = 8,
#| fig.height = 10

par(mfrow=c(1,2))
for (i in 1:length(site_ID)) {
  filt = climateData.ET.HargreavesSamani$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.HargreavesSamani$temporal$year,
              climateData.ET.HargreavesSamani$temporal$month,
              climateData.ET.HargreavesSamani$temporal$day)
  plot(d[filt], climateData.ET.HargreavesSamani$temporal$et[filt],
              col='black',lty=1, xlim = c(ISOdate(2010,1,1), ISOdate(2010,12,1)),
              ylim=c(0, 30),type='l', ylab='ET [mm/d]',xlab='Date')

  filt = climateData.ET.JensenHaise$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.JensenHaise$temporal$year,
              climateData.ET.JensenHaise$temporal$month,
              climateData.ET.JensenHaise$temporal$day)
  lines(d[filt], climateData.ET.JensenHaise$temporal$et[filt],
              col='red',lty=1)

  filt = climateData.ET.Makkink$temporal$Location.ID==407214
  d = ISOdate(climateData.ET.Makkink$temporal$year,
              climateData.ET.Makkink$temporal$month,
              climateData.ET.Makkink$temporal$day)
  lines(d[filt], climateData.ET.Makkink$temporal$et[filt],
              col='green',lty=1)

  filt = climateData.ET.McGuinnessBordne$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.McGuinnessBordne$temporal$year,
              climateData.ET.McGuinnessBordne$temporal$month,
              climateData.ET.McGuinnessBordne$temporal$day)
  lines(d[filt], climateData.ET.McGuinnessBordne$temporal$et[filt],
              col='blue',lty=1)

  filt = climateData.ET.MortonCRAE.potentialET$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.MortonCRAE.potentialET$temporal$year,
              climateData.ET.MortonCRAE.potentialET$temporal$month,
              climateData.ET.MortonCRAE.potentialET$temporal$day)
  lines(d[filt], climateData.ET.MortonCRAE.potentialET$temporal$et[filt],
              col='black',lty=2)

  filt = climateData.ET.MortonCRAE.wetarealET$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.MortonCRAE.wetarealET$temporal$year,
              climateData.ET.MortonCRAE.wetarealET$temporal$month,
              climateData.ET.MortonCRAE.wetarealET$temporal$day)
  lines(d[filt], climateData.ET.MortonCRAE.wetarealET$temporal$et[filt],
              col='red',lty=2)

  filt = climateData.ET.MortonCRAE.actualarealET$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.MortonCRAE.actualarealET$temporal$year,
              climateData.ET.MortonCRAE.actualarealET$temporal$month,
              climateData.ET.MortonCRAE.actualarealET$temporal$day)
  lines(d[filt], climateData.ET.MortonCRAE.actualarealET$temporal$et[filt],
              col='green',lty=2)

  filt = climateData.ET.MortonCRWE$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.MortonCRWE$temporal$year,
              climateData.ET.MortonCRWE$temporal$month,
              climateData.ET.MortonCRWE$temporal$day)
  lines(d[filt], climateData.ET.MortonCRWE$temporal$et[filt],
              col='blue',lty=2)

  filt = climateData.ET.MortonCRWE.shallowLake$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.MortonCRWE.shallowLake$temporal$year,
              climateData.ET.MortonCRWE.shallowLake$temporal$month,
              climateData.ET.MortonCRWE.shallowLake$temporal$day)
  lines(d[filt], climateData.ET.MortonCRWE.shallowLake$temporal$et[filt],
              col='black',lty=3)

  filt = climateData.ET.Turc$temporal$Location.ID == site_ID[i]
  d = ISOdate(climateData.ET.Turc$temporal$year,
              climateData.ET.Turc$temporal$month,
              climateData.ET.Turc$temporal$day)
  lines(d[filt], climateData.ET.Turc$temporal$et[filt],
              col='red',lty=3)

}

legend(x='topright', legend=c(
  'Hargreaves Samani (ref. crop)',
  'Jensen Haise (PET)',
  'Makkink (ref. crop)',
  'McGuinness Bordne (PET)',
  'Morton CRAE (PET)',
  'Morton CRAE (wet areal ET)',
  'Morton CRAE (actual areal ET)',
  'Morton CRWE (PET)',
  'Morton CRWE (shallow Lake)',
  'Turc (ref. crop, non-humid)'),
  lty = c(1,1,1,1,2,2,2,2,3,3),
  col=c('black','red','green','blue','black','red','green','blue','black','red')
)
```
