---
title: "4. Area weighted evapotranspiration estimates"
description: >
  This vignette shows how to estimate various measures of the area weighted evapotranspiration.
output:
  html_document:
    code_folding: show
  rmarkdown:
    html_vignette
vignette: >
  %\VignetteIndexEntry{4. Area weighted evapotranspiration estimates}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

```{r setup}
library(BOMcatchr)
```

This example calculates and plot various estimates of evaportranspiration. Ten different estimates of area weighted evapotranspiration over one year at catchment 407214 (Victoria, Australia) are derived.

## Make netCDF files
Like the other vignettes, the netCDF data grids need to be built.

First, let's define the start and end dates for data grids and the file names.

```{r}
date.from = as.Date("2010-01-01","%Y-%m-%d")
date.to = as.Date("2010-12-31","%Y-%m-%d")

ncdfFilename = tempfile(fileext='.nc')
```

Next, let's make the data grids over this period.
```{r}
status = grid_build(ncdfFilename = ncdfFilename,
                   updateFrom = date.from,
                   updateTo = date.to,
                   vars = c('precip','tmin', 'tmax',
                   'vprp_3pm', 'solarrad'))
```

## Load catchment boundaries
Now that we have the meteorological data we can begin extracting data for two catchments. To make it easy to get the require location data, the package includes the function \code{extract_locations}, which pulls vector spatial data from the BOM Australian Hydrological Geospatial Fabric.  Here it to used to obtain the boundaries for two catchments within the Loddon basin, Victoria. However, shape files of polygons can easily be used once imported into R.
```{r}
catch <- extract_locations('catchment',c('407214','407220'))
site_ID <- catch$stationno
```

## Extract daily precipitation and PET data
Next, the 11 different measures of evapotranspiration that can be derived from the available gridded data are calculated for 12 months.

The estimation of ET uses the _evapotranspiration_ package. It requires a set of constants, which are loaded as follows.
```{r}
data(constants,package='Evapotranspiration')
```

Next, all 10 ET measures are derived. The first step is to build a list of the inputs for each derived variable and then collate them into a list of lists. Note, the list of inputs must start with the function name and then the named inputs - excluding the extraction time steps, the required gridded observation variable names, the coordinates and the land surface elevation of the grid cells.

Also, the inputs \code{ET_missing_method} and \code{ET_abnormal_method} normally do not need to be defined, but are done so here because the short extraction length of less than two years prohibits the use of the default daily average, \code{DoY average}.

Lastly, here only built-in derived functions are used. However, user defined functions can also be input.
```{R}
derived_var_fns <- list(ET_HS = list( FUN = 'ET_HargreavesSamani',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'daily'),
                        ET_JH = list( FUN = 'ET_JensenHaise',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'daily'),
                        ET_M =  list( FUN = 'ET_Makkink',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'daily'),
                        ET_MB = list( FUN = 'ET_McGuinnessBordne',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'daily'),
                        ET_T =  list( FUN = 'ET_Turc',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'daily'),
                        ET_CRAE_PET = list( FUN = 'ET_MortonCRAE',
                                            ET_Mortons_est = 'potential ET',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'monthly'),
                        ET_CRAE_aPET = list(FUN = 'ET_MortonCRAE',
                                            ET_Mortons_est = 'wet areal ET',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'monthly'),
                        ET_CRAE_aET = list( FUN = 'ET_MortonCRAE',
                                            ET_Mortons_est = 'actual areal ET',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'monthly'),
                        ET_CRWE_lake = list(FUN = 'ET_MortonCRWE',
                                            ET_Mortons_est = 'shallow lake ET',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'monthly'),
                        ET_CRWE_PET  = list(FUN = 'ET_MortonCRWE',
                                            ET_Mortons_est = 'potential ET',
                                            ET_missing_method = 'neighbouring average',
                                            ET_abnormal_method = 'neighbouring average',
                                            ET_constants = constants,
                                            ET_timestep = 'monthly')
                        )
```
Now we can call the extraction functiion and derive all ET estimates in one call.
```{R}
climateData = extract_data(ncdfFilename=ncdfFilename,
                               extractFrom = date.from,
                               extractTo = date.to,
                               vars = c('precip', 'tmin', 'tmax', 'vprp_3pm','solarrad'),
                               vars_derived = derived_var_fns,
                               locations = catch,
                               temporal.timestep = 'daily',
                               temporal.fn.inner ='sum',
                               spatial.fn = 'IQR')

```
Next let's plot each measure of the area weighted evaporation at each catchment.

```{r class.source = 'fold-hide'}
#| fig.width = 8,
#| fig.height = 10

par(mfrow=c(2,1))
for (i in 1:length(site_ID)) {
  filt = climateData$temporal$Location.ID == site_ID[i]

  d = ISOdate(climateData$temporal$year,
              climateData$temporal$month,
              climateData$temporal$day)
  plot(d[filt],
       climateData$temporal$ET_HS[filt],
       col='black',
       lty=1,
       xlim = c(ISOdate(2010,1,1),
       ISOdate(2010,12,1)),
       ylim=c(0, 30),
       type='l',
       ylab='ET [mm/d]',
       xlab='Date',
       main=paste('Catchment ID',site_ID[i])
       )

  lines(d[filt],
        climateData$temporal$ET_JH[filt],
        col='red',
        lty=1)

  lines(d[filt],
        climateData$temporal$ET_M[filt],
        col='green',
        lty=1)

  lines(d[filt],
        climateData$temporal$ET_MB[filt],
        col='blue',
        lty=1)

  lines(d[filt], climateData$temporal$ET_T[filt],
        col='purple',
        lty=3)

  lines(d[filt],
        climateData$temporal$ET_CRAE_PET[filt],
        col='black',
        lty=2)

  lines(d[filt],
        climateData$temporal$ET_CRAE_aPET[filt],
        col='red',
        lty=2)

  lines(d[filt],
        climateData$temporal$ET_CRAE_aET[filt],
        col='green',
        lty=2)

  lines(d[filt],
        climateData$temporal$ET_CRWE_lake[filt],
        col='blue',
        lty=2)

  lines(d[filt],
        climateData$temporal$ET_CRWE_PET[filt],
        col='black',
        lty=3)

}

legend(x='topright', legend=c(
  'Hargreaves Samani (ref. crop)',
  'Jensen Haise (PET)',
  'Makkink (ref. crop)',
  'McGuinness Bordne (PET)',
  'Turc (ref. crop, non-humid)',
  'Morton CRAE (PET)',
  'Morton CRAE (wet areal ET)',
  'Morton CRAE (actual areal ET)',
  'Morton CRWE (PET)',
  'Morton CRWE (shallow Lake)'),
  lty = c(1,1,1,1,1,2,2,2,3,3),
  col=c('black','red','green','blue','purple','black','red','green','blue','black')
)
```
